Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2024-02-28 paper/markdown/supplementary_material.Rmd
Revised figures, tables, manuscript and supplement

For distribution testing (normality, variance, Gini index), the packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA), _rstatix_ [@Kassambara2021] and _DescTools_ [@Signorell2022] were utilized. 
Consistency of psychometric tools was assessed with the packages _psych_ [@Revelle2022], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Clustering tendency was investigated with _factoextra_ [@Kassambara2020] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and _rstatix_ [@Kassambara2021].

@@

The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
anxiety (GAD-7 and GAD-2: 7- and 2-item general anxiety disorder scale) [@Spitzer2006; @Junker2023], 
depression (PHQ: patient health questionnaire, PHQ-9, PHQ-2 and PHQ-8) [@Lowe2002; @Grafe2004; @Kroenke2002a; @Junker2023], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
common persistent somatic symptoms as a substitute for somatization (PHQ-15) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016]. 
In the study survey, items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').

Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
For a comparison of frequency of anxious and depressive symptoms in the study cohort with the estimates for the German population [@Junker2023], presence of anxious and depressive symptoms was considered for GAD-2 $\geq$ 3 and PHQ-2 $\geq$ 3 points, respectively. 
For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].

@@

Additional information on study variables and their stratification schemes are presented in __Supplementary Table S\@ref(tab:tab-analysis-vars)__.

@@

Monthly estimates of population frequency of symptoms of anxiety and depression in the general German population during the study survey (`r ptsd$dataset$survey_date %>% range %>% paste(collapse = ' to ')`) were derived from the [data repository of the German mental health monitoring](https://github.com/robert-koch-institut/Hochfrequente_Mental_Health_Surveillance) [@Junker2023]. 
Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a nation-wide Austrian microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].

@@

Frequency of traumatic events, manifest PTSD and PTSD symptoms in the current cohort was compared with original reports providing estimates of frequency of traumatic events and PTSD in the general population [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017] and in a cohort of Swiss mountain rescuers [@Mikutta2022]. 
Frequencies of symptoms of anxiety and depression measured with the GAD-2 and PHQ-2 instruments, respectively, were compared with estimates for the general German population gauged by the German mental health monitoring during the study survey [@Junker2023]. 
Frequency and severity of depressive symptoms measured by the GAD-8 tools in the study cohort [@Kroenke2002a] were compared with the Austrian microcensus estimates. 
In those analyses, bootstrap with B = 1000 resamples was used for computation of the bias-corrected and accelerated (BCA) 95% confidence intervals (function `bmap()`, package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat)). 
Significant differences were assumed for non-overlapping 95% confidence intervals of the mental symptom frequency estimates.

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by 
its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares), 
good separability between the clusters measured by mean silhouette width [@Rousseeuw1987], 
and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `var()`, `silhouette()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020]. 
By this means, three mental health clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).

Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$frac_var[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$sil_width[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$frac_var[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$sil_width[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

@@

To investigate effects observations excluded from the analysis due to an incomplete psychometric data set on definition of the mental health clusters, the excluded observations with less than 80% of missing psychometric entries were subjected to imputation with the 9-nearest neighbor algorithm implemented by the `impute.knn()` function from the R package _impute_ [@Hastie2023]. 
Subsequently, the mental health clusters were developed for the analysis cohort and for the analysis cohort appended with the imputed observations. 
Analogically, to explore effects of prior mountain sport accidents, the mental health clusters were developed for the entire cohort, the individuals having experienced a mountain accident in the past, and participants without prior mountain sport accidents. 
To evaluate effects of annual household income, the mental clusters were developed for the analysis cohort and the analysis cohort without, respectively, participants with no income, yearly income < 30000 Euro, yearly income ranging from 30000 to 45000 Euro and yearly income > 45000 Euro. 
To assess the impact of injury severity, the mental clusters were developed for the analysis cohort and the analysis cohort depleted of, respectively, individuals with mild (AIS 1), moderate (AIS 2), and severe-to-critical injuries (AIS $\geq$ 3). 
To investigate the effect of hospitalization, the mental health clusters were developed separately for the analysis cohort, non-hospitalized and hospitalized participants. 
To assess the impact of accident year, the mental clusters were developed for the analysis cohort without, respectively, accidents in 2018, 2019 and 2022. 
In each case, the unsupervised PAM/cosine distance clustering algorithm was used, i.e. the procedure employed for definition of the genuine mental health clusters. 
The effects of extending or depletion of the data set on the quality of the clustering structure was evaluated by comparing fractions of the explained clustering variance between the clustering solutions for the particular data set and for the analysis cohort [@Thabane2013]. 
Additionally, distribution of sizes of the neutral, PTG and PTS clusters expressed as percentage of all clustered observations was compared (__Supplementary Figure S\@ref(fig:fig-cluster-se)__ and __S\@ref(fig:fig-cluster-miss)__).

@@

flextable::flextable(suppl_tables$no_response) %>% 
  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
   footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury scale; ICD-10: 10th version of the International Statistical Classification of Diseases and Related Health Problems.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

__Supplementary Figure \@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__

_Mountain sport types (A), accident months (B) and age class distribution (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023. Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety. In the comparison of sport types, non-sportive mountain accidents (e.g. during hunting or forestry) were excluded and sport types in the study cohort re-classified according to the scheme of the Austrian Board of Mountain Safety. In the comparison of the age class distribution, only individuals older than 20 years were included. Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic. Pairwise differences for particular sport types, months and age classes were assessed by $\chi^2$ test corrected for multiple testing with the Holm method. Percentages of categories within the complete observation sets are presented as bar plots. Bars are labeled with the corresponding percent values. Total observation numbers are displayed in the plot legends. Global effect sizes and p values are presented in the plot captions. Significant results of post-hoc tests are highlighted with bold X axis labels._

_alpine skiing: alpine skiing and snowboarding in a ski resort and lift accidents; ski touring: ski touring and freeride skiing; air: air sports and paragliding; water: water sports._

@@

_(A) Comparison of frequency of anxious and depressive symptoms measured by the GAD-2 and PHQ-2 instruments in the study cohort with frequency of anxious and depressive symptoms obtained for the general German population by monthly mental health monitoring._

@@

_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta at al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_

@@

_(A) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._

_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study data set visualized as a heat map. Average cross-distances with 95% percentile range are indicated in the tiles._

@@

_Distribution of employment status, frequency of mountain sport accidents in the past, distribution of the accidents in the seasons, accident sport types, frequency of being alone during the accident and number of injured persons in the mental health clusters. Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Percentages of variable's categories in the entire study cohort are presented in stack plots. Effect sizes and p-values are displayed in the plot captions. Numbers of observations in the clusters are presented in the X axes._

>>>>

For distribution testing (normality, variance, Gini index), the packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA), _rstatix_ [@Kassambara2021] and _DescTools_ [@Signorell2022] were utilized. 
Consistency of psychometric tools was assessed with the packages _psych_ [@Revelle2022], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Clustering tendency was investigated with _factoextra_ [@Kassambara2020] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and _rstatix_ [@Kassambara2021]. 
Bootstrapped estimates of prevalence of mental disorder symptoms and traumatic events with their 95% confidence intervals were computed with the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).

@@

The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9 and PHQ-8) [@Lowe2002; @Grafe2004; @Kroenke2002a], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
common persistent somatic symptoms as a substitute for somatization (PHQ-15) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016]. 
In the study survey, items of the PTGI and PTSD instruments, and questions concerning flashbacks were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').

Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 10 [@Humer2022], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].

@@

Accident sport types were classified in accordance with the annual accident report of the Austrian Board of Mountain Safety [@OsterreichischesKuratoriumfurAlpineSicherheit2024] as follows: 
_alpine skiing/snowboarding_ (all types of alpine ski and snowboard sports in secured areas of ski resorts: pistes, secured slopes, lifts and cable cars), 
_ski touring/freeride_ (all types of alpine ski and snowboard sports performed in non-secured areas of ski reports, e.g. ski touring, freeride), 
_cross-country skiing_, 
_sledding_ (all types of sledding sports such as sledding and bob-sledding), 
_ice climbing_ (ice and mixed climbing in the winter season), 
_hiking_, 
_climbing_ (sport and rock climbing, bouldering), 
_mountaineering_ (high mountain and glacier tours), 
_biking_ (all types of biking including road, cross-country and mountain biking), 
_air sport_ (only paragliding present in our cohort), 
_water sport_, and 
_other_ (non-classifiable sport types such as ski jumping). 
Alpine skiing/snowboarding, ski touring/freeride, cross-country skiing, sledding, and ice climbing were classified as winter sports. 
Hiking, climbing, mountaineering, biking, air sport, and water sport were classified as summer sports. 
Additional information on study variables and their stratification schemes are presented in __Supplementary Table S\@ref(tab:tab-analysis-vars)__.

@@

Data on frequency of clinically relevant anxious (GAD-7 $\geq$ 10) and depressive (PHQ-9 $\geq$ 10) symptoms in Austria in 2022 were obtained from a publication by Humer et al. [@Humer2022]. 
Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a pre-pandemic nation-wide Austrian 2019 microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].

@@

Frequency of traumatic events, manifest PTSD and PTSD symptoms in the current cohort was compared with original reports providing estimates of frequency of traumatic events and PTSD in the general population [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017] and in a cohort of Swiss mountain rescuers [@Mikutta2022]. 
Frequencies of clinically relevant symptoms of anxiety and depression measured, respectively, with GAD-7 and PHQ-9, were compared with a representative sample of the general Austrian population reported by Humer et al. [@Humer2022]. 
Frequency and severity of depressive symptoms measured by the PHQ-8 tool in the study cohort [@Kroenke2002a] were compared with the Austrian 2019 microcensus estimates [@StatistikAustria2019a]. 
In those analyses, bootstrap with B = 1000 resamples was used for computation of the bias-corrected and accelerated (BCA) 95% confidence intervals (function `bmap()`, package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat)). 
Significant differences were assumed for non-overlapping 95% confidence intervals of the mental symptom frequency estimates.

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by 
its good explanatory performance measured by the fraction of explained clustering variance (ratio of the total between-cluster sum of squares to the total sum of squares), 
good separability between the clusters measured by mean silhouette width [@Rousseeuw1987], 
and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `summary()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
The number of clusters was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020]. 
By this means, three mental health clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).

Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Additionally, the reproducibility of the clustering structure was assessed by 
mean silhouette width as a measure of general cluster separation (training: `r signif(semi_clust$stats$sil_width[1], 2)`, test: `r signif(semi_clust$stats$sil_width[2], 2)`) [@Rousseeuw1987], 
fraction of observations with negative silhouette widths as a metric of misclassification rate (training: `r signif(semi_clust$stats$frac_misclassified[1], 2)`, test: `r signif(semi_clust$stats$frac_misclassified[2], 2)`) [@Rousseeuw1987], 
fraction of explained clustering variance defined as ratio of the total between-cluster sum of squares to the total sum of squares (training: `r signif(semi_clust$stats$frac_var[1], 2)`, test: `r signif(semi_clust$stats$frac_var[2], 2)`), 
and fraction of 5-nearest neighbors placed in the same cluster as a measure of neighborhood preservation (training: `r signif(semi_clust$stats$frac_np[1], 2)`, test: `r signif(semi_clust$stats$frac_np[2], 2)`) [@Venna2001]. 
These statistics were computed with the `summary()` method from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
The results of the comparison of the numeric statistics are presented in __Supplementary Figure S\@ref(fig:fig-cosine)A__. 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)B__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)C__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

@@

To investigate effects observations excluded from the analysis due to an incomplete psychometric data set on definition of the mental health clusters, the excluded observations with less than 80% of missing psychometric entries were subjected to imputation with the 9-nearest neighbor algorithm implemented by the `impute.knn()` function from the R package _impute_ [@Hastie2023]. 
Subsequently, the mental health clusters were developed for the analysis cohort and for the analysis cohort appended with the imputed observations. 
Analogically, to explore effects of prior mountain sport accidents, the mental health clusters were developed for the entire cohort, the individuals having experienced a mountain accident in the past, and participants without prior mountain sport accidents. 
To evaluate effects of annual household income, the mental clusters were developed for the analysis cohort and the analysis cohort without, respectively, participants with no income, yearly income < 30000 Euro, yearly income ranging from 30000 to 45000 Euro and yearly income > 45000 Euro. 
To assess the impact of injury severity, the mental clusters were developed for the analysis cohort and the analysis cohort depleted of, respectively, individuals with mild (AIS 1), moderate (AIS 2), and severe-to-critical injuries (AIS $\geq$ 3). 
To investigate the effect of hospitalization, the mental health clusters were developed separately for the analysis cohort, non-hospitalized and hospitalized participants. 
To assess the impact of accident year, the mental clusters were developed for the analysis cohort without, respectively, accidents in 2018, 2019 and 2022. 
In each case, the unsupervised PAM/cosine distance clustering algorithm was used, i.e. the procedure employed for definition of the genuine mental health clusters. 
The effects of extending or depletion of the data set on the quality of the clustering structure was evaluated by comparing fractions of the explained clustering variance between the clustering solutions for the particular data set and for the analysis cohort [@Thabane2013]. 
Additionally, sizes of the neutral, PTG and PTS clusters expressed as percentage of all clustered observations were compared (__Supplementary Figure S\@ref(fig:fig-cluster-se)__ and __S\@ref(fig:fig-cluster-miss)__).

@@

flextable::flextable(suppl_tables$no_response) %>% 
  set_widths(c(2.8, 6.3, 6.3, 2.5, 2.2)) %>% 
   footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury scale; ICD-10: 10th version of the International Statistical Classification of Diseases and Related Health Problems.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

__Supplementary Figure S\@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__

_Mountain sport types (A), accident months (B) and age class distribution (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023. Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety. In the comparison of sport types, non-sportive mountain accidents (e.g. during hunting or forestry) were excluded from the national statistic. Non-classifiable sport types (category 'other') in the study cohort were excluded from the analysis presented in (C). In the comparison of the age class distribution, only individuals older than 20 years were included. Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic. Pairwise differences for particular sport types, months and age classes were assessed by $\chi^2$ test corrected for multiple testing with the Holm method. Percentages of categories within the complete observation sets are presented as bar plots. Bars are labeled with the corresponding percent values. Total observation numbers are displayed in the plot legends. Global effect sizes and p values are presented in the plot captions. Significant results of post-hoc tests are highlighted with bold X axis labels._

_alpine skiing: alpine skiing and snowboarding in secured areas of ski resorts; ski touring/freeride: alpine skiing and snowboarding performed besides secured areas of ski resorts; sledding: sledding and bob sledding; ice climbing: ice and mixed climbing; climbing: bouldering, sport and rock climbing; mountaineering: high mountain and glacier tours; biking: road, cross-country and mountain biking._

@@

_(A) Comparison of frequency of anxious and depressive symptoms measured by the GAD-7 and PHQ-9 instruments in the study cohort with frequency of anxious and depressive symptoms reported in a representative sample of Austrian population in 2022 by Humer and colleagues._

@@

_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_

@@

_(A) Comparison of numeric statistics of quality of clustering structures between the training and test subset of the study cohort. Mean silhouette width served as a measure of cluster separation. Fractions of observations with negative values of the silhouette width statistics were used to gauge misclassification rates. Explained clustering variance was defined as ratio of the total between-cluster sum of squares to the total sum of squares. Neighborhood preservation was measured by a fraction of the 5-nearest neighbors placed in the same cluster. SS: sum of squares, NN: nearest neighbors._

_(B) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._

_(C) Cosine cross-distances between the mental health clusters in the training and test subsets of the study data set visualized as a heat map. Average cross-distances with 95% percentile range are indicated in the tiles._

@@

_Distribution of employment status, frequency of mountain sport accidents in the past, distribution of the accidents in the seasons, frequency of being alone during the accident and number of injured persons in the mental health clusters. Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Percentages of variable's categories in the entire study cohort are presented in stack plots. Effect sizes and p-values are displayed in the plot captions. Numbers of observations in the clusters are presented in the X axes._

```{r fig-cluster-sport, fig.width = figur::convert(suppl_figures$cluster_sport, to = 'in')$w, fig.height = figur::convert(suppl_figures$cluster_sport, to = 'in')$h, fig.cap = 'Mountain sport types in the clusters.'}

suppl_figures$cluster_sport$plot

__Supplementary Figure S\@ref(fig:fig-cluster-sport). Mountain sport types in the clusters.__

_The global difference in distribution of accident sport types between the mental health clusters was compared by $\chi^2$ test with Cramer's V effect size statistic. Post-hoc testing for differences in frequency of particular sport types between the clusters was done with Holm-corrected $\chi^2$ test with Cramer's V effect size statistic. Percentages of sport types within the clusters are presented in bar plots. Effect size statistic and p value for the global difference are indicated in the plot caption. Results of post-hoc tests are displayed in the Y axis. Numbers of observations in the clusters are shown in the plot legend._
